---
title: "faux"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{faux}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(faux)
library(patchwork)
```


## Simulate data for a factorial design

### sim_design

See the [Simulate by Design vignette](articles/sim_design.html) for more details.

```{r}
#| label: fig-sim-design
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Default design plot"
between <- list(pet = c(cat = "Cat Owners", 
                        dog = "Dog Owners"))
within <- list(time = c("morning", 
                        "noon", 
                        "evening", 
                        "night"))
mu <- data.frame(
  cat    = c(10, 12, 14, 16),
  dog    = c(10, 15, 20, 25),
  row.names = within$time
)
df <- sim_design(within, between, 
                 n = 100, mu = mu, 
                 sd = 5, r = .5)
```


```{r}
#| label: fig-design
#| fig-cap: "Plot the data with different visualisations."
p1 <- plot_design(df)
p2 <- plot_design(df, "pet", "time")

p1 + p2 + plot_layout(ncol = 1)
```

### pipeline

You can also use the new pipeline functions to simulate data from a fixed design.

```{r}
pipe_df <- add_random(id = 200) |>
  add_between(pet = c(cat = "Cat Owners", 
                      dog = "Dog Owners")) |>
  add_within(time = c("morning", 
                      "noon", 
                      "evening", 
                      "night")) |>
  add_dv(vars = "dv",
         between = list(pet = "cat"),
         mean = c(10, 12, 14, 16),
         sd = 5, r = .5, 
         empirical = TRUE) |>
  add_dv(vars = "dv",
         between = list(pet = "dog"),
         mean = c(10, 15, 20, 25),
         sd = 5, r = 0, 
         empirical = TRUE) |>
  add_labels(id = "ID", 
             dv = "Score", 
             pet = "Pet Type", 
             time = "Time of Day")
  
```

```{r}
sample_params(pipe_df)
```


```{r}
pipe_wide <- long2wide(pipe_df)
# check IDs
```



## Simulate new data from an existing data table

See the [Simulate from Existing Data vignette](articles/sim_df.html) for more details.

```{r sim-df}
new_iris <- sim_df(iris, 50, between = "Species") 
```

```{r}
#| label: fig-iris-sim
#| echo: false
#| fig-cap: "Simulated iris dataset"

new_iris |>
  dplyr::mutate_if(is.numeric, round, 1) |>
  dplyr::mutate(data = "Simulated") |>
  dplyr::bind_rows(dplyr::mutate(iris, data = "Original")) |>
  ggplot2::ggplot(ggplot2::aes(x = Sepal.Width, 
                               y = Sepal.Length, 
                               color = Species)) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(method = "lm", formula = "y~x") +
  ggplot2::facet_wrap(~data)
```

## Simulate data for a mixed design

You can build up a cross-classified or nested mixed effects design using piped functions. See the [contrasts vignette](articles/contrasts.html) for more details.

```{r}
# simulate 20 classes with 20 to 30 students per class
data <- add_random(class = 20) |>
  add_random(student = sample(20:30, 20, replace = TRUE), 
             .nested_in = "class") |>
  add_between(.by = "class", 
              school_type = c("private","public"), 
              .prob = c(5, 15)) |>
  add_between(.by = "student",
              gender = c("M", "F", "NB"),
              .prob = c(.49, .49, .02))
```

```{r}
#| echo: false
dplyr::count(data, school_type, gender) |>
  knitr::kable()
```