---
title: "Simulate Correlated Variables"
author: "Lisa DeBruine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate Correlated Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(8675309)
```

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(faux)
```

The `rnorm_multi()` function makes multiple normally distributed vectors with specified parameters and relationships. 

## Quick example

For example, the following creates a sample that has 100 observations of 3 variables, drawn from a population where A has a mean of 0 and SD of 1, while B and C have means of 20 and SDs of 5. A correlates with B and C with r = 0.5, and B and C correlate with r = 0.25.


```{r multirnorm-example}

dat <- rnorm_multi(n = 100, 
                  mu = c(0, 20, 20),
                  sd = c(1, 5, 5),
                  r = c(0.5, 0.5, 0.25), 
                  varnames = c("A", "B", "C"),
                  empirical = FALSE)

```


`r sample_params(dat) %>% knitr::kable()`
Table: Sample stats

### Specify correlations {#spec_r}

You can specify the correlations in one of four ways:

* A single r for all pairs
* A vars by vars matrix
* A vars\*vars length vector
* A vars\*(vars-1)/2 length vector

#### One Number

If you want all the pairs to have the same correlation, just specify a single number. 

```{r}
bvn <- rnorm_multi(100, 5, 0, 1, .3, varnames = letters[1:5])
```

`r sample_params(bvn) %>% knitr::kable()`
Table: Sample stats from a single rho

#### Matrix

If you already have a correlation matrix, such as the output of `cor()`, you can specify the simulated data with that.

```{r vvmatrix}
cmat <- cor(iris[,1:4])
bvn <- rnorm_multi(100, 4, 0, 1, cmat, 
                  varnames = colnames(cmat))
```

`r sample_params(bvn) %>% knitr::kable()`
Table: Sample stats from a correlation matrix

#### Vector (vars\*vars)

You can specify your correlation matrix by hand as a vars\*vars length vector, which will include the correlations of 1 down the diagonal.

```{r}
cmat <- c(1, .3, .5,
          .3, 1, 0,
          .5, 0, 1)
bvn <- rnorm_multi(100, 3, 0, 1, cmat, 
                  varnames = c("first", "second", "third"))
```

`r sample_params(bvn) %>% knitr::kable()`
Table: Sample stats from a vars\*vars vector

#### Vector (vars\*(vars-1)/2)

You can specify your correlation matrix by hand as a vars\*(vars-1)/2 length vector, skipping the diagonal and lower left duplicate values.

```{r}
rho1_2 <- .3
rho1_3 <- .5
rho1_4 <- .5
rho2_3 <- .2
rho2_4 <- 0
rho3_4 <- -.3
cmat <- c(rho1_2, rho1_3, rho1_4, rho2_3, rho2_4, rho3_4)
bvn <- rnorm_multi(100, 4, 0, 1, cmat, 
                  varnames = letters[1:4])
```

`r sample_params(bvn) %>% knitr::kable()`
Table: Sample stats from a (vars\*(vars-1)/2) vector

### empirical

If you want your samples to have the *exact* correlations, means, and SDs you entered, set `empirical` to TRUE.

```{r}
bvn <- rnorm_multi(100, 5, 0, 1, .3, 
                  varnames = letters[1:5], 
                  empirical = T)
```

`r sample_params(bvn) %>% knitr::kable()`
Table: Sample stats with empirical = TRUE

## Pre-existing variables

Us `rnorm_pre()` to create a vector with a specified correlation to one or more pre-existing variables. The following code creates a new column called `B` with a mean of 10, SD of 2 and a correlation of r = 0.5 to the `A` column.

```{r}
dat <- rnorm_multi(varnames = "A") %>%
  mutate(B = rnorm_pre(A, mu = 10, sd = 2, r = 0.5))
```

```{r, echo = FALSE, results='asis'}
sample_params(dat) %>% knitr::kable(digits = 3)
```

Set `empirical = TRUE` to return a vector with the **exact** specified parameters.

```{r}
dat$C <- rnorm_pre(dat$A, mu = 10, sd = 2, r = 0.5, empirical = TRUE)
```

```{r, echo = FALSE, results='asis'}
sample_params(dat) %>% knitr::kable(digits = 3)
```


You can also specify correlations to more than one vector by setting the first argument to a data frame containing only the continuous columns and r to the correlation with each column.

```{r}
dat$D <- rnorm_pre(dat, r = c(.1, .2, .3), empirical = TRUE)
```

```{r, echo = FALSE, results='asis'}
sample_params(dat) %>% knitr::kable(digits = 3)
```

Not all correlation patterns are possible, so you'll get an error message if the correlations you ask for are impossible.

```{r, error = TRUE, perl = FALSE}
dat$E <- rnorm_pre(dat, r = .9)
```
